2020 U.S. Presidential Election Forecast

### 2024-03-15 paper/paper.qmd
introduction

Education is also an interesting factor. In the past, non-college white voters used to support Democrats while college-educated white voters supported Republicans (@Harris). However, there has been a switch in this trend as 61 percent of non-college white voters showed their support wheres just 45 percent of college-educated white voters did in the exit polls (@Harris). Only 37 percent of those without a degree cast their votes for Democrats while 53 percent with a degree did so (@Harris). We categorized education into four categories: 'High school or less', 'Some college', 'College degree', 'Postgrad'.

Race also needs some attention because normally non-white groups are highly in favour of Democrats regardless of candidates and white swing by depending on candidates. According to the statistics collected in 2016, 93% of black, 71% of Latino, 68% of Asian support democrats while only 41% of white support democrats (@Prokop). As white voters make up 74% of the voting population, it is really important for both parties to attain this demographic group.

@@

IPUMS ("Integrated Public Use Microdata Series") is a website that offers database of samples of the American population from the American Community Surveys of 2000-present. These samples provide rich qualitative information on the long-term changes in the population. We selected the '2019 ACS' data (@citeIPUMS) as the post-stratification dataset for our research. The ACS is an ongoing survey that collects data monthly, which is then combined into 1-year, 3-year, and 5-year aggregates. It then uses stratified sampling where the U.S population is broken down into sub-groups and initial weights are assigned to each respondent.

@@

The key steps in MRP involve initially selecting a dataset for model development. In this case, we utilized survey data from the Voter Study Group (@citeSurveyData). The next step is to construct a model with this smaller dataset; here, we employed logistic regression based on the survey data, formulated as seen in equation \ref{eq:logit}. Following model creation, it is then applied to a broader dataset to estimate population characteristics. For our analysis, Census data from IPUMS (@citeIPUMS) served as this larger dataset.

To predict an individual's likelihood of voting for Donald Trump, we aim to construct a logistic regression model leveraging data from the Voter Study Group (@citeSurveyData) and applying post-stratification with Census Data (@citeIPUMS). Given that logistic regression is suited for binary outcomes, we've introduced a variable, 'consider_trump', which assigns a 1 if the respondent indicates a plan to vote for Donald Trump, and a 0 for intentions to vote for other candidates, with 0 encompassing both "No" and "Other" responses.

@@

Once the logistic regression model is developed, we'll use the `predict()` function in R (@citeR) to apply our model to Census data (@citeIPUMS), breaking down the dataset into categories based on sex, race, age group, education level, and state. This will give us the likelihood of individuals within each category voting for Donald Trump. These predictions allow us to analyze potential outcomes like the popular vote winner or the electoral college vote distribution.

@@

@tbl-model-coefficients shows the estimates for the coefficients that will fit into our logistic regression equation. These coefficients will fit into Equation \ref{eq:logit}, and were calculated using data from the Voter Study Group (@citeSurveyData). The table is made using `kable` from `knitr` (@citeKnitr) and is formatted using `kableExtra` (@citekableExtra).

We use the `stan_glm()` function in R (@citeR) for our regression analysis, specifically because we are dealing with a binary outcome: whether a voter supports Donald Trump or not. The nature of our data suggests an S-shaped distribution rather than a linear one, making logistic regression a better fit than linear regression. This approach is advantageous, especially when paired with post-stratification, as it allows us to better represent under-represented groups in our analysis. For instance, despite having only 7 responses from Alaska in our survey data (@citeSurveyData), through multilevel regression and post-stratification, we can adjust this to effectively represent over 4500 individuals.

@@

@fig-coefficient-estimates presents the coefficients derived from logistic regression on the survey data (@citeSurveyData). It also includes error bars, indicating the confidence interval for each coefficient estimate. In interpreting these coefficients, it's essential to understand that positive values suggest a greater likelihood of voting for Donald Trump, whereas negative values indicate a tendency to vote for other candidates, such as Joe Biden.

@@

From @fig-distribution-by-state, it seems that the majority of the states support Biden. Only 7 states out of 51 have its point estimate greater than 50 percent for Trump. The horizontal lines of confidence intervals of some states overlapping the green mark give some hope for the Republicans. Excluding these contesting states, however, our model suggests that only 3 states are definitely in favor of Trump whereas 35 states are definitely supporting Biden.

@@

From @fig-distribution-by-state, we can observe that Trump leads in mid-west of America, Montana (MT), Wyoming (WY), Mississippi (MS), Maine (ME), Missouri (MO), Kentucky (KY), and Utah (UT). Even the competing states like Louisiana (LA), New Mexico (NM), Tennessee (TN) are also mid-west part of the country, which are rural areas with low population density. This is correctly estimated since in the real world as Trump won in these states: MT, WY, MS, MA, MI, KY, UT. He also won in LA and TN, which we considered were competing states. We can go more in depth of this study with the benefit of hindsight. We estimated that the biggest victory for Trump would be Montana (MT) with just below 70%. In reality, he won by 59.6%, which is lower than we thought but still sit within our error values. The next state with the highest proportion of support for Trump was Wyoming (WY) from our model with 65%, which is close to the election poll result of 69%. We were very successful with Mississippi (MS) with an estimation of 57% and the poll result of 57.5%. Maine (ME) was another very close estimate with 52% and the poll result of 53.1%. 56.8% voted for Trump in Missouri (MO), which was slightly higher than our estimation of 51%. We estimated that only 26% of voters in Hawaii (HI) would vote for Trump with error margins of +-10%, which matched 34.3% for Trump in the election. 41.9% for Trump in Colorado (CO) lie within our estimate of 37% with +-5% margins.

There were some cases where our estimation was incorrect in terms of who won the state but the results lie within our error margins. For instance, we estimated Utah (UT) in favor of the Democrats with a slight margin of 51% for Biden but Biden only received 37.6% in the election. This still lies within our error values that we estimated to be from 43% to 58%. However, there were some errors in our estimation where our model failed to predict the Trump winning in the number of states in general. Our model predicted Louisiana (LA) for Biden with 51% but he only received 39.9% of the votes. This value is lower than our error margins. 62% of the residents in Alabama, a Republican stronghold in the last eleven presidential elections had voted for Trump while our model estimated 45% of votes for Trump. The republicans won in Florida (FL) with a slight margin of 51.2% but we predicted an estimation of 45% with +-2% margins. What we estimated to be the strongest Democrat state was North Dakota (ND) with an estimation of only 23% with +-12% margins but in reality 65.1% of the residents voted for Trump. In general, big cities and large suburban areas along the west coast and the east coast favored Biden whereas Trump supporters are mostly in the rural areas in the mid-south and mid-north states from the election results.

>>>>

Education is also an interesting factor. In the past, non-college white voters used to support Democrats while college-educated white voters supported Republicans [@Harris]. However, there has been a switch in this trend as 61 percent of non-college white voters showed their support wheres just 45 percent of college-educated white voters did in the exit polls [@Harris]. Only 37 percent of those without a degree cast their votes for Democrats while 53 percent with a degree did so [@Harris]. We categorized education into four categories: 'High school or less', 'Some college', 'College degree', 'Postgrad'.

Race also needs some attention because normally non-white groups are highly in favour of Democrats regardless of candidates and white swing by depending on candidates. According to the statistics collected in 2016, 93% of black, 71% of Latino, 68% of Asian support democrats while only 41% of white support democrats [@Prokop]. As white voters make up 74% of the voting population, it is really important for both parties to attain this demographic group.

@@

IPUMS ("Integrated Public Use Microdata Series") is a website that offers database of samples of the American population from the American Community Surveys of 2000-present. These samples provide rich qualitative information on the long-term changes in the population. We selected the '2019 ACS' data [@citeIPUMS] as the post-stratification dataset for our research. The ACS is an ongoing survey that collects data monthly, which is then combined into 1-year, 3-year, and 5-year aggregates. It then uses stratified sampling where the U.S population is broken down into sub-groups and initial weights are assigned to each respondent.

@@

The key steps in MRP involve initially selecting a dataset for model development. In this case, we utilized survey data from the Voter Study Group [@citeSurveyData]. The next step is to construct a model with this smaller dataset; here, we employed logistic regression based on the survey data, formulated as seen in equation \ref{eq:logit}. Following model creation, it is then applied to a broader dataset to estimate population characteristics. For our analysis, Census data from IPUMS [@citeIPUMS] served as this larger dataset.

To predict an individual's likelihood of voting for Donald Trump, we aim to construct a logistic regression model leveraging data from the Voter Study Group [@citeSurveyData] and applying post-stratification with Census Data [@citeIPUMS]. Given that logistic regression is suited for binary outcomes, we've introduced a variable, 'consider_trump', which assigns a 1 if the respondent indicates a plan to vote for Donald Trump, and a 0 for intentions to vote for other candidates, with 0 encompassing both "No" and "Other" responses.

@@

Once the logistic regression model is developed, we'll use the `predict()` function in R [@citeR] to apply our model to Census data [@citeIPUMS], breaking down the dataset into categories based on sex, race, age group, education level, and state. This will give us the likelihood of individuals within each category voting for Donald Trump. These predictions allow us to analyze potential outcomes like the popular vote winner or the electoral college vote distribution.

@@

@tbl-model-coefficients shows the estimates for the coefficients that will fit into our logistic regression equation. These coefficients will fit into Equation \ref{eq:logit}, and were calculated using data from the Voter Study Group [@citeSurveyData]. The table is made using `kable` from `knitr` [@citeKnitr] and is formatted using `kableExtra` [@citekableExtra].

We use the `stan_glm()` function in R [@citeR] for our regression analysis, specifically because we are dealing with a binary outcome: whether a voter supports Donald Trump or not. The nature of our data suggests an S-shaped distribution rather than a linear one, making logistic regression a better fit than linear regression. This approach is advantageous, especially when paired with post-stratification, as it allows us to better represent under-represented groups in our analysis. For instance, despite having only 7 responses from Alaska in our survey data [@citeSurveyData], through multilevel regression and post-stratification, we can adjust this to effectively represent over 4500 individuals.

@@

@fig-coefficient-estimates presents the coefficients derived from logistic regression on the survey data [@citeSurveyData]. It also includes error bars, indicating the confidence interval for each coefficient estimate. In interpreting these coefficients, it's essential to understand that positive values suggest a greater likelihood of voting for Donald Trump, whereas negative values indicate a tendency to vote for other candidates, such as Joe Biden.

@@

From @fig-distribution-by-state, it seems that the majority of the states support Biden. Only 7 states out of 51 have its point estimate greater than 50% for Trump. The horizontal lines of confidence intervals of some states overlapping the green mark give some hope for the Republicans. Excluding these contesting states, however, our model suggests that only 3 states are definitely in favor of Trump whereas 35 states are definitely supporting Biden.

@@

From @fig-distribution-by-state, we can observe that Trump leads in mid-west of America, Montana (MT), Wyoming (WY), Mississippi (MS), Maine (ME), Missouri (MO), Kentucky (KY), and Utah (UT). Even the competing states like Louisiana (LA), New Mexico (NM), Tennessee (TN) are also mid-west part of the country, which are rural areas with low population density. This is correctly estimated since in the real world as Trump won in these states: MT, WY, MS, MA, MI, KY, UT. He also won in LA and TN, which we considered were competing states. We can go more in depth of this study with the benefit of hindsight. We estimated that the biggest victory for Trump would be Montana (MT) with just below 70%. In reality, he won by 59.6%, which is lower than we thought but still sit within our error values. The next state with the highest proportion of support for Trump was Wyoming (WY) from our model with 65%, which is close to the election poll result of 69%. We were very successful with Mississippi (MS) with an estimation of 57% and the poll result of 57.5%. Maine (ME) was another very close estimate with 52% and the poll result of 53.1%. 56.8% voted for Trump in Missouri (MO), which was slightly higher than our estimation of 51%. We estimated that only 26% of voters in Hawaii (HI) would vote for Trump with error margins of ±10%, which matched 34.3% for Trump in the election. 41.9% for Trump in Colorado (CO) lie within our estimate of 37% with ±5% margins.

There were some cases where our estimation was incorrect in terms of who won the state but the results lie within our error margins. For instance, we estimated Utah (UT) in favor of the Democrats with a slight margin of 51% for Biden but Biden only received 37.6% in the election. This still lies within our error values that we estimated to be from 43% to 58%. However, there were some errors in our estimation where our model failed to predict the Trump winning in the number of states in general. Our model predicted Louisiana (LA) for Biden with 51% but he only received 39.9% of the votes. This value is lower than our error margins. 62% of the residents in Alabama, a Republican stronghold in the last eleven presidential elections had voted for Trump while our model estimated 45% of votes for Trump. The republicans won in Florida (FL) with a slight margin of 51.2% but we predicted an estimation of 45% with ±2% margins. What we estimated to be the strongest Democrat state was North Dakota (ND) with an estimation of only 23% with ±12% margins but in reality 65.1% of the residents voted for Trump. In general, big cities and large suburban areas along the west coast and the east coast favored Biden whereas Trump supporters are mostly in the rural areas in the mid-south and mid-north states from the election results.
